---
title: "Intro to R - Learning Resources"
output:
  html_document: 
    theme: united
    highlight: tango
    toc: yes
    toc_float: yes
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("./Images/LATree.PNG"),
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:200px;')
```


# Learning Resources.{-}

We thank you for attending the Introduction R course and we hope you  enjoyed and have now a basic understanding of the language and are able to do some further learning independently.

We have compiled a list of free resources to further independent learning.

## RStudio Tip and Tricks

| Title       | Description |
| ----------- | ----------- |
| [RStudio Keyboard Shortcuts](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts )| help to run commands faster|



## Books and Sites

| Title              | Description|
| -----------        | -----------|
| [Cheatsheets](https://www.rstudio.com/resources/cheatsheets/ ) | A cheat sheet covers basic commands with minimalist explanations. It expects users to have some basic familiarity with R/R Studio. It includes data importing, manipulation, statistics, and plotting etc    |
|  [Rseek.org](https://rseek.org/ )             | RSeek is a custom search engine that can help you find information on the official website, the CRAN, the archives of the mailing lists, the documentation of R and even selected websites. It is more effective than a simple Google search.  |
| [R-Bloggers](https://www.r-bloggers.com/)              | R-Bloggers.com is a blog aggregator of content collected from bloggers who write about R. It has an effective search tool that allows searching in all articles published by the contributing bloggers.   |
|  [R for Data Science](https://r4ds.had.co.nz/)          |This book will teach you how to do data science with R,covers the	tidyverse,ggplot2,dplyr,tidyr,readr     |
|[R Graphics Cookbook](https://r-graphics.org/)            | 	A practical guide that provides more than 150 recipes to help you generate high-quality graphs quickly	ggplot2,data visualization |
|  [Mordern Drive](https://moderndive.com/1-getting-started.html)             |This book assumes no prerequisites: no algebra, no calculus, and no prior programming/coding experience. This is intended to be a gentle introduction to the practice of analyzing data and answering questions using data the way data scientists, statisticians, data journalists, and other researchers would, covers data visualisation, importing data, linear regression, data manipulation    |
|  [ONS R coding standards](https://datasciencecampus.github.io/coding-standards/r.html)          |provides guidance on what style to use when programming and what tools we use (and recommend) to enable you to make projects more accessible and easier to share or collaborate on.   |
| [An Intro to R](https://cran.r-project.org/doc/manuals/R-intro.html#Preface)     | This introduction to R is derived from an original set of notes describing the S and S-PLUS environments written in 1990–2 by Bill Venables and David M. Smith when at the University of Adelaide.       |
| [R Language Definition](https://cran.r-project.org/doc/manuals/r-release/R-lang.html)   | R is a system for statistical computation and graphics. It provides, among other things, a programming language, high level graphics, interfaces to other languages and debugging facilities. This manual details and defines the R language.The purpose of this manual is to document the language per se. That is, the objects that it works on, and the details of the expression evaluation process, which are useful to know when programming R functions. Major subsystems for specific tasks, such as graphics, are only briefly described in this manual and will be documented separately.     |
|[R In a Nutshell](https://paulvanderlaken.files.wordpress.com/2017/08/r_in_a_nutshell.pdf)             | This book is designed to be a concise guide to R. It’s not intended to be a book about statistics or an exhaustive guide to R            |
| [A (very) short introduction to R](https://cran.r-project.org/doc/contrib/Torfs+Brauer-Short-R-Intro.pdf )      | Really short and quick overview of R           |
|[Introduction to R](https://ramnathv.github.io/pycon2014-r/)   | These are notes for an introductory R workshop  for Python Programmers.        |



## Other Courses

This list contains other topics we currently do not cover on the course.

| Title       | Description |
| ----------- | ----------- |
| [SimpleR](https://cran.r-project.org/doc/contrib/Verzani-SimpleR.pdf )      |Using R for Introductory Statistics        |



## Online Courses

| Title      | Description |
| ----------- | ----------- |
| [Swirl](https://swirlstats.com/students.html)      | The swirl R package makes it fun and easy to learn R programming and data science. If you are new to R, have no fear.      |

