---
title: "Intro to R - Session Two"
output:
 html_document: 
    theme: united
    highlight: tango
    toc: yes
    toc_float: yes
    number_sections: true
---
```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("./Images/LATree.PNG"),
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:200px;')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Session Two.

***

## Learning Objectives
**The goal of this session is to:**

* Be able to create vectors to store data.
* Understand the indexing system in R.
* Be able to select elements from vectors.
* Understand how Vectors Operate.
* Understand the difference between Data frames vs Tibbles.


## Recap on Vectors

* All elements in a vector must have the same type/class
* They are created using the  `c()` function, which stands for combine.
* Elements within the vector are separated by `commas`

You will be dealing a lot with vectors so we will go through some exercises on how to operate with them.
    
### Vectors Practice

In your script leave yourself a comment to remind that this exercise is about vectors.

#### Creating Vectors {.tabset .tabset-fade}

##### Exercise

```{r, eval = F}

#Exercise 2 -Vectors


```


Use the combine function to create some Vectors called,

`days_of_the_week <-`

`numbers <-`

##### Answers

```{r, eval = F}
#Exercise 2 -Vectors

days_of_the_week <- c("Monday","Tuesday",
                      "Wednesday","Thursday",
                      "Friday","Saturday",
                      "Sunday")

numbers <- 1:20


```

## Vector Operations {.tabset .tabset-fade .tabset-pills}

Now create 2 new vectors using arithmetic on the numbers Vector

### Exercise
```{r}

#even_numbers <- 


#odd_numbers <- 


```


### Answers

```{r echo=TRUE}

numbers <- 1:20

even_numbers <- numbers*2

odd_numbers <- numbers*2 - 1

```

## Vectors Operations 2 {.tabset .tabset-fade .tabset-pills}

### Exercise
```{r}

# What does the code below do?

vector1 <- c(4:10)

mean(vector1)

#Create a Vector which subtracts the mean of vector1 from each element of vector1.

#Minus_mean <- 
 

```
 
 
 

### Answers

```{r echo=TRUE}

vector1 <- c(4:10)

mean(vector1)


minus_mean <- vector1 - mean(vector1)

minus_mean

```




## Vectors Operations 3 {.tabset .tabset-fade .tabset-pills}

As R is a vectorised language, this means operations are applied to each element of the vector automatically, without the need to loop through the vector.

### Exercise
```{r}
#Create a vector assigned to 'a' with the numbers 1 to 5

#a <- 

#Make another vector assignred to 'b' with the numbers 1 to 3
  
#b <- 
  
#Add 'a' and 'b' together

#a + b

#Can you work out what has happened?

#Try again with 'a' as the numbers 1 to 6.
  
  
 
 

```
 
 

### Answers

```{r echo=TRUE}

a <- 1:5

b <- 1:3

a+b

# This gives the following warning:
# [1] 2 4 6 5 7
# Warning message:
#   In a + b : longer object length is 
# not a multiple of shorter object length

a <- 1:6

a+b

# This doesn't give a warning.
# We need to be careful that objects 
# have the same number of items.

```
All of the operations we have seen above were performed in an **element-wise** manner. This means the operation was performed on each respective element from each vector. In R this is implemented with a lot of flexibiliy. e.g. We could perform operations with vectors of different sizes. Other languages may not have the same flexibility and will give you an error.



## Vectors Indexing {.tabset .tabset-fade .tabset-pills}

Selecting elements from a vector, Notice that R indexes from 1 (not 0 like other languages).

We select elements with `[ ]`

### Exercise
```{r}

# try selecting the 3rd element from days_of_the_week

#days_of_the_week[]

#find out how to select muiltiple elements from a vector, can you select;

#The week days

# Tuesday and Thursday
  
  

```
 
 

### Answers

```{r echo=TRUE}

days_of_the_week <- c("Monday","Tuesday",
                      "Wednesday","Thursday",
                      "Friday","Saturday",
                      "Sunday")


# 3rd day
days_of_the_week[3]
# week days
days_of_the_week[1:5]
# Tuesday and Thursday
days_of_the_week[c(2,4)] 
  # Here we create a selection 
  # vector inside the [].

```

## Data Frames and Tibbles {.tabset .tabset-fade .tabset-pills}

**Recap on Data Frames**

* Can hold variables of different types
* Columns are variables 
* Rows are instances

In the next session we are going to use special type of data frames called **tibbles.**

They are becoming a standard way to work with data, they are very useful when working with large data sets.
 
**The key benefits of tibbles are;**

* The class of each column is displayed above it.

* The dimensions of the tibble are shown at the top.

* Prints the first 10 rows by default.

* You can only view columns that fit on the screen.


### Tibble

for more information type `vignette("tibble")` in your script.

```{r}

library(tibble)

as_tibble(iris)


```

### Data Frame
```{r}

iris
  
```


