---
title: "Intro to R - Session Two"
output:
 html_document: 
    theme: united
    highlight: tango
    toc: yes
    toc_float: yes
    number_sections: true
---
```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("./Images/LATree.PNG"),
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:200px;')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Session Two.

***

## Learning Objectives
**The goal of this session is to:**

* Be able to create vectors to store data.
* Understand the indexing system in R.
* Be able to select elements from vectors.
* Understand how Vectors Operate.


***

## Recap on Vectors

* All elements in a vector must have the same type/class
* They are created using the  `c()` function, which stands for combine.
* Elements within the vector are separated by `commas`

You will be dealing a lot with vectors so we will go through some exercises on how to operate with them.

***
## Vectors

In the previous section we looked at data types, where we said;

```{r}

a <- 10

```
Suppose you wanted to store more than one value for 10. e.g values between 10-50

```{r, eval=FALSE}

a <- 10, 20, 30, 40, 50

a <- (10, 20, 30,40,50)


```

### The c() Function {.tabset .tabset-fade}

The code above will not work, which is where the c() funtion comes in. 

#### Question

Use the help section in R to find out what the c() function does.

```{r, eval=TRUE}

?c()

```

#### Show Answer
```{r}

?c()

#or

help(c)

```
The c() function is used for creating a vector in R.

### Creating Vectors {.tabset .tabset-fade}

There are several other ways of creating a vector:

You can create a vector by using the the assignement operator.
or you can use the sequence function seq().

We can also use the **:** to create a vector of consecutive numbers.

```{r}

my_first_vector <- c(1,2,3,4,5)

my_first_vector

my_second_vector <- seq(2,6, by=2)

my_second_vector

my_third_vector <- 6:10

my_third_vector

```
#### **Question**

In your script leave yourself a comment to remind that this exercise is about vectors.

Create 2 vectors, one numeric and one character, e.g favourate_movies and lucky_numbers. (Feel free to be creative here)

Now combine the 2 vectors to create a new vector.

Find out what kind of vector your new vector is.

```{r, eval=FALSE}

favourate_movies <- c("Frozen", "The Lion King", "Moana") 
  
lucky_numbers <- c(7, 12, 15)   

```

#### **Show Answer**

We can see below that our new vector is a character vector, as vectors are supposed to be of the same data type, R will convert the data of the 2 vectors to be the same, this is known as **coercion.**

```{r}
favourate_movies <- c("Frozen", "The Lion King", "Moana") 
  
lucky_numbers <- c(7, 12, 15)   

movies_and_numbers <- c(favourate_movies,lucky_numbers)

typeof(movies_and_numbers)

```


## Vector Operations {.tabset .tabset-fade}

As we know vector in R is a data element so we can perform arithmetic operations on vector in R such as addition, subtraction and multiplication.

As R is a vectorised language, this means operations are applied to each element of the vector automatically, without the need to loop through the vector.

This enables us to perfom calculations much faster.

```{r}

vector_addition <- my_first_vector + my_third_vector

vector_multiplication <- my_first_vector * my_third_vector


```

### **Question**

Try and add my_first_vector and my_second_vector.

Can you figure out what has happened?



```{r, eval=FALSE}

vector_after_addition <- my_first_vector + my_second_vector

vector_after_addition  

```

### **Show Answer**

When we try to add vectors of different length the R gives a warning message as the longer object length is not a multiple of the shorter object length.

As we can see R has recycled the elements in the shorter vector.


```{r}
vector_after_addition <- my_first_vector + my_second_vector

vector_after_addition  

   
```
## Vector Indexing{.tabset .tabset-fade}

Elements in a vector can be selected using `[ ]`.

Note that R indexes from 1 (not 0 like other languages).

As an example, going back to my favourate movies, to pick the second item in the list, I would simply, use the square bracketS and reference **2** as shown below.

```{r}

favourate_movies[2] 


```



### **Question**

Create another vector containing days of the week (Monday - Sunday).

Select the third day of the week.

Select the week days.

Select Tuesday and Thursday.

```{r, eval=FALSE}

days_of_the_week <- 
  

```

### **Show Answer**


```{r}
days_of_the_week <- c("Monday","Tuesday",
                      "Wednesday","Thursday",
                      "Friday","Saturday",
                      "Sunday")


# 3rd day
days_of_the_week[3]


# week days
days_of_the_week[1:5]

#Tuesday and Thursday

days_of_the_week[c(2,4)] 
  # Here we create a selection 
  # vector inside the [].
```

