---
title: "Intro to R - Session Six"
output:
  html_document: 
    theme: united
    highlight: tango
    toc: yes
    toc_float: yes
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("./Images/LATree.PNG"),
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:200px;')
```


# SESSION SIX.

## LEARNING OBJECTIVES

**The goal of this session is to:**

* Have an understanding of what packages are.
* Be able to load and install a package.
* Be able to check package versions and R version.
* Be able to import data from multiple formarts.
* Be able to inspect loaded data and select elements within the dataframe.
* Be able to export data.


## CONDITIONAL STATEMENTS.

A conditional statement will check if a condition is TRUE or FALSE.

So if our data meets this criteria apply this action or function otherwise do nothing.

Useful if we want a function to run only data means a certain condition, this is termed `control flow`.

We use comparison and logical operators.

### Comparison Operators

| Operator | Description | 
| ------ |:------:| 
|   ==   | equal to | 
|  !=    | not equal to |   
| <      | less than|
| >      | greater than |
| <=     | less than or eaqual to|
| >=     | greater than or equal to|

### Logical Operators

| Operator| Description| 
| ------  |:------:| 
|   &     | And | 
|   \|    | Or |
|    !    | Not|


You might want an action to be perfomed depending on the outcome, to be able to do this in R, we use the;

- if statement

- else statement

- elseif statement

### IF statement

They follow the basic structure below,


```{r, eval=FALSE}

If (logical condition){
  
Commands
  
}

```

- So, `if` followed by round brackets, in which we enter our logical condition 

- then curly brackets in which we enter a series of commands. Its good practice to put the commands on a new line as its easier to read, notice the code is also indented so we can see that its part of the if statement.

`If statements` on their own do not work with vectorised arguments. They will only work with a single element and not a vector of elements. If applied to a vector it will only check the first obersevation, if this doesnt meet the condtion then no statements will be executed, you will get an error/warning.

#### Example

We can look at the example below, lets say we set our `height` as 6.

Our condition is `height is equal to 6`

Our Command is if `height is equal to 6` to `print me too`

```{r}

height <- 6

if(height == 6){
  print("me too!")
}

```
### ELSE Statement

It has to be used together with an `if statement`.

The `else statement` does not require a condition; the code of an else statement gets executed when ever the result of an `if statement` is False.

They follow the basic structure below,

```{r, eval=FALSE}

If (logical condition){
  
Commands
  
}else{New command
  }

```

All we have added is else and an additional commands.
Else has to be on the same line as the closing bracket or else wont work.

#### Example

Going back to our previous example in which we said `height <-6`.

We can add an else statement.

This will get executed if the `height` is not equal to 6.

So lets say `height` is now equal to 7.

The first condition is FALSE (height ==6 ), hence it executes the else statement.


```{r}

height <- 7

if(height == 6){
  print("me too!")
}else{
  print("Not my height")
  }

```
### ELSE IF

This allows you to further customise your control flow. With `else if`, you can add as many conditions as we want.

The else if statement specifies a new condition if the first condition is false.

Keep in mind that R ignores the remainder of the control structure once a condition has been found that is TRUE and the corresponding expressions have been executed.

They follow the basic structure below,

it’s important that the else if keyword comes on the same line as the closing bracket of the previous part of the control structure.

```{r, eval=FALSE}

If (logical condition){
  
Commands- executed if logical condition is true
  
}else if( second condition) {
  commands- executed if the logical condition is false and second condtion is true
}else if (third condition){
  commands- executed if logical condition and second condition are false
} else {
  commands- executed when all conditions are false
}

```

#### Example
```{r}

height <- 9

if (height == 6){
  print("me too!, my height is 6")
  
}else if( height == 7) { print(
  "me too!, my height is 7")
 
}else if (height == 8){print(
  "me too!, my height is 8")
  
}else {
  print("Not my height")
  
}

```

#### Exercise{.tabset .tabset-fade}

##### **Question**

Now lets try and practise.

can you write an `else if` statement, that looks at `grades`

It should class grades 

- between 0 and 40 as failed.

- between 41 and 50 a D

- between 51 and 60 as C

- between 61 and 70 as B

- between 71 and 100 as an A

It should also print the grade and an optional comment on the grade.

##### **Show Answer**

```{r}
grade <- c(56)

if (grade <= 40 & grade >=0){
  print("You have absolutely Failed!!")
  print("We are really sorry for you")
}else if (grade <= 50 & grade >=41){
  print("You have got a D")
  print("Need to work harder")
}else if (grade <= 60 & grade >=51){
  print("You have got a C")
  print("Not bad, can do better")
}else if(grade <= 70 & grade >=61){
  print("You have got a B")
  print("Well done")
}else if (grade >= 71 & grade <=100){
  print("You have got an A")
  print("Congratulations, this is excellent")
} else {
  print("Error, grades can not be less than 0 or more than 100")
  }


```


## LOOPS

A loop is a way to repeat a sequence of instructions under certain conditions. 

They allow you to automate parts of your code that are in need of repetition.

In R you have multiple options when repeating calculations, `vectorized operations`, `for loops`, and `apply functions`.

We covered 'vectorized operations` in session 2 (section 1.4), we will look at loops in session and we will not cover the apply functions.

### Types of loops
 
 We are going to look at `for loops` and `while loops`.

- For loops

- While loops



#### For loops {.tabset .tabset-fade}

These are the most common types.

They follow the basic structure below,

```{r, eval=FALSE}

for (Variable in Vector){
  commands
}


```
 
- So we start with the word `for`

- followed by `()` brackets where we specify an index variable, `i` is commonly used but it can be anything that you want.

- then the word `in`

- then specify a vector

- followed by `{}` curly brackets which will have our commands which could be multiple lines of code.

lets look at an example.

Lets start by creating a vector, I have created one with **even numbers from 2 to 10**.


```{r}

example_vector <- c( 2,4,6,8,10 )

```

We will now run our `for loop` on our vector.

```{r}


for( i in example_vector ){
  print( i + 1 )
}

```

So our index variable is `i`, our vector is `example_vector` and then our command is to add `1`.

So the first time the loop runs **(Iteration)** the index variable `i` was assigned to our first number in our vector `2` and since our command is add a 1, we get a result of 3.

An **iteration** is a single run through lines of code contained in a loop

So the second time the loop runs the index variable `i` is assigned to the second number in our vector `4` and this adds a 1, we get a result of 5.

This will continue until the command has been run into all elements of the vector.

we cant really use our put like this as it just prints to the console, if we wanted to save our output into something we wanted to use. We could assign it.

```{r, eval=FALSE}

for( i in example_vector ){
  odd_numbers <- append(odd_numbers, i + 1 )
}

odd_numbers

```

The code above will not work as odd_numbers does not exist outside the brackets, to solve this issue we create an empty vector, called `odd_numbers` and we add our results into in using the `append function`.

**append() function simply adds elements to a vector.**.

```{r}

odd_numbers <- c()

for( i in example_vector ){
  odd_numbers <- append(odd_numbers, i + 1 )
}

odd_numbers

```

##### **Question**

##### **Show Answer**

#### While loops{.tabset .tabset-fade}

The main difference between the while loops and the for loops is that we dont specify the number of iterations in the loop.

We want the loop to keep running until a specified condition is met.

They follow the basic structure below,

```{r, eval=FALSE}

while (logical condition){
  commands
}


```
- So, stat with the word `while`

- followed by the `()` brackets in which we specify our logical condition, using the **conditional operators**.

- and then `{}`, which hold our commands. These commands will be run repeatedly in order of appearance while the specified condtion is TRUE.

lets look at an example,

- If we say `x` is equal to 2.

- and we set out condtion to be `x < 6`

- then print `x`

- and then `add a 1`

```{r}

x <- 2

while(x<6){
  print(x)
  x <- x+1
}


```

- so we know that starting value is `2`. ( As set by `x <- 2` ).

- the first thing the loop does it to test it against the condition `x<6`

- in our case x is `2` which is less than `6`, so the condtition is `TRUE`

- So we can move on to the next part of the loop, which is `print x` which is equal to `2`. (This is our first 2 in the output)

- The next step is to override the value of x by adding a 1. **(x <- x+1 )**

- As our new x value is still less than 6, the condition is TRUE, so it will print x which is now 3 and get a new value for x (x <- x+1), which will now be 4.

- This continues like this until we get to x is 5, this passes the fist conditional statement `x<6` so it will print a 5 but after we add the 1 `x <- x+1 ` and x becomes 6, it is not less than 6, so it fails the first conditional statement and the loop can not continue, hence our results are between 2-5.


**The order of the commands is important.**

##### **Question**

- Can you work out what the ouput would be if we reversed the order of the commands?. i.e, start with the  `y <- y+1` and then `print x`

- What would happen if we removed the line of code `y <- y+1`

```{r, eval=FALSE}

y <- 2

while(y<6){
  y <- y+1
  print(y)
}


```


##### **Show Answer**

```{r}

y <- 2

while(y<6){
  y <- y+1
  print(y)
}


```
- As we can see we now get values between 3 and 5.

- So we start we with y is equal to 2, which passes the first condition `y<6`

- We then add a 1 to it `y <- y+1`, so 2 + 1. Which is equal to 3.

- Our fist output is 3

- And that continues on.

-For the next question, if we removed the code `y <- y+1` y would always be less than 6 and you will be stuck in an infinite loop, as R will keep running the code. You will have to hit the stop sign to exit the loop.

#### Exercise loops{.tabset .tabset-fade}



##### **Question**

Code a while loop with the following characteristics:

- The condition of the `while` loop, should check if the temperature is hiegher than 20

- The first command in the body should print out `"Its way too hot!"`

- The next command should decrease the `temperature` by 2 and assign this new value to `temperature` again.

##### **Show Answer**

```{r}
temperature <- 40

# Code the while loop
while (temperature > 20) {
  print("It way too hot!")
  temperature <- temperature - 2
}

# Print out the temperature variable
temperature

```

##### **Question**

We can combile a `loop` and and `if else statement`.

create a vector, with values from 1-7 and assign it to number_of_takeaways_ordered 

create a for loop with an if and else statement shows that if you order more than 2 takeaways a week it print too many and if you order less than two it prints enough.

##### **Show Answer**

```{r}

number_of_takeaways_ordered <-c(1,2,3,4,5,6,7)

for (i in number_of_takeaways_ordered){
  if (i > 3){
    print("too many")
  }else{
  print("enough")
  }
  print(i)
}

```

