---
title: "Extended Exercise 4.2: Working with Dataframes 1"
output:
  html_document: 
    theme: united
    highlight: textmate
    toc: yes
    toc_float: yes
    number_sections: true
---

```{r, echo=FALSE, eval=FALSE}
htmltools::img(src = knitr::image_uri("./Images/LATree.PNG"),
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:200px;')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
```

In this exercise, we will practice the data frame operations that we have
covered so far (and introducea few more tips and tricks along the way); the
different steps below are set up so that they build increasingly complex
operations. To remind yourself of the core material we covered see
<mark>XXX</mark> and <mark>XXX</mark>.
For the exercises presented here, we will use extracts from published data on
the UK CPI by the Office for National Statistics. The source data used, and more
information on the background to it, can be found
[here](https://www.ons.gov.uk/economy/inflationandpriceindices/datasets/consumerpriceinflation).\
Please note that:
-   The example data used here are prepared specifically for the purposes of
these exercises, and do not represent official statistics in any way.
-   While these exercises are designed to cover the kind of operations that may
be used in real-world CPI production workflows, they are not intended as
directly representing true CPI calculations.

# Reading, viewing, and cleaning up data {.tabset}

We will start by working with two example data sets in XLSX format. The first is a table of CPI indices between 2008 and 2021 (`Data\CPI indices example.xlsx`). The second is a list of CPI codes and corresponding category names (`Data\CPI codes and names.xlsx`). Two start off with -
1. Load each of these datasets into suitable R objects (dataframes or tibbles).
2. Perform some initial checks of each dataset, ensuring that the data has been read in correctly.

##  {.unnumbered}

## Hint {.unnumbered}

Tabset hint text

## Example Answer {.unnumbered}

Example answer text

```{r code_block_1}
```

# Conclusion {.unnumbered}

We have now created a dataset within R, performed an initial exploration of it before saving it to `.xlsx` to share with colleagues.
