---
title: "Extended Exercise - Manipulating Strings"
output:
  html_document: 
    theme: united
    highlight: textmate
    toc: yes
    toc_float: yes
    number_sections: true
---

```{r, echo=FALSE, eval=FALSE}
htmltools::img(src = knitr::image_uri("./Images/LATree.PNG"),
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:200px;')
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The aim for this exercise is to familiarise ourselves with common methods of
standardising text data. This makes it easier to then summarise and compare
the data. This tutorial will cover lowercasing, removing extra spaces, how to
replace words and how to split strings.

We will first run through these techniques on a (fairly contrived) dataset
before giving you the chance to practise on some "real" data.

This first dataset contains a sample of sales data from a juice shop. 
Unfortunately the barcodes aren't working so the staff have to type in the name
of the products they've sold, so there are several inconsistencies. We want to
write some code that counts how many of each type of juice have been sold.

First lets read in the data and take a look at it.

```{r read_juice}
juice_sales <- read.csv('../Data/juice_sales.csv')

juice_sales
```



## Plan

-> first look in xlsx to get a feel for the data and what steps might need to be taken

-> lowercase everything (option to uppercase)

-> removing spaces from start and end

-> single spaces between words

-> replacing synonyms

-> splitting on comma? (when two+ are on one line)


exercise: do the same/similar with cause of death data
```{r read death data}
cause_of_death = read.csv('../Data/cause_of_death.csv')
```