---
title: "Extended Exercise 4.1: Building Dataframes"
output:
  html_document: 
    theme: united
    highlight: textmate
    toc: yes
    toc_float: yes
    number_sections: true
---

```{r, echo=FALSE, eval=FALSE}
htmltools::img(src = knitr::image_uri("./Images/LATree.PNG"),
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:200px;')
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Although we will usually use R to read in data from other sources (`.csv` or
`.xlsx` for example), this extended exercise illustrates how we can create
data within an R script itself.

In this exercise we will create and explore a hypothetical 'basket of goods'
dataset. Throughout, we will provide example answers but encourage you to come
up with your own examples.

# Create a vector of 10 items that might be included {.tabset}
A basket of goods dataset will contain average prices for items and goods
bought by a large proportion of the population. Let's first create a vector of
some of these items. Aim for at least 10 from a variety of categories. Save
this vector to a well named object!

## {-}

## Hint {-}
The `c()` function can be used to create a vector of objects.

## Example Answer {-}
```{r item_vector}
basket_of_goods_items <-
  c('Bread Rolls', 'Crumpets', 'Pasta', 'Haggis', 'Tea bags', 'Whisky', 'Beer',
    'Shirt', 'Kilt', 'Jacket')
```
A [kilt](https://en.wikipedia.org/wiki/Kilt) is part of the national dress in
Scotland.

# Create vector of the prices of these items {.tabset}
Next we want to create another vector containing the prices of each of these
items. Don't worry too much about the actual prices here.

It is important however that these two vectors are the same length, so that
each item has a corresponding price, and vice versa. Can you check this
programmatically?

## {-}

## Hint {-}
Again, we can use the `c()` function to create a vector.

## Example Answer {-}
```{r price_vector}
basket_of_goods_price <-
  c(1.20, 0.85, 0.39, 2.6, 3.5, 45, 12, 30, 400, 100)
```

# Check datatypes
# Create dataframe with appropriate column names
# Perform summaries on created df
# Create another one (from a different year/region)
# Join together in list
# save to xlsx (as seperate sheets(?))