---
title: "Intro to R - Extra Material 1.1 - How to get help"
output:
  html_document: 
    theme: united
    highlight: textmate
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
    number_sections: true
---

```{r, echo=FALSE, eval=FALSE}
htmltools::img(src = knitr::image_uri("./Images/LATree.PNG"),
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:200px;')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this appendix to the Introduction to R course, we provide some more detailed
suggestions on how to find help or solve errors/bugs in your R code; it is an
expansion of Section 5 in [Chapter
1](../Course_content/Chapter_1_Getting_Started_with_R.html) of the main course.

# Finding a solution for a specific problem
When you start coding in any language, you feel like you need to memorise
everything to be able to code. However, this is not how most programmers do it:
they use Google (or other search engines) to find the a solution to the parts of
their problem.

An important thing to note: **When searching for R-related questions use use "R
cran"[^1] instead of just "R". Using just the letter "r" is not very specific for
obvious reasons.**

[^1]: CRAN is the official online repository for R downloads and packages.

## Error messages
When using programs and functions you will encounter error messages. One
important thing to remember is that when something breaks, the **first error
message is the most/only important one**. This is because the first message is
the first point at which the problem happened, all subsequent errors are likely
to be because the first part broke.

Make sure you read the error message thoroughly, it may just be caused by a
syntax error (missing bracket, semi-colon, etc) which has a very easy fix.  

When you get an error that you don't understand, google the exact error message,
except for any parts which are machine specific (like the path in your home
directory where the error occurred). It's important to remember error messages
exist somewhere in a codebase, it's not just the program making conversation
with you, so try and find them in the documentation/forums.

## Forums and documentation
### Documentation
Documentation for functions and packages are where you can find all the
specifications that program/function can do. Each function available in R,
whether that is in the "base" installation or within packages you load
externally, will have documentation included with it. You can access this in R
through the methods described in Section 5.2 of [Chapter
1](../Course_content/Chapter_1_Getting_Started_with_R.html).  

It is important to remember that these are written to be precise and extensive
and so can be a bit daunting when starting out. However, if you want to go
deeper and understand all the options and arguments of a function, the
documentation should have all the information you need.

### Forums
As highlighted in Section 5.3 of [Chapter
1](../Course_content/Chapter_1_Getting_Started_with_R.html), online forums such
as [StackOverflow](https://stackoverflow.com/) are a great place to find
solutions to common problems. One of the most difficult parts of finding the
solution to your problem is how to search for it in the first place. Finding the
specific words that define your problem will save you a lot of looking through
answers in forums.

For example: if you searched for a "vehicle fixer" would not give you the
specific results searching for "mechanic" would.

Another key element is using Google keywords that will focus the search engine
and help it give you more relevant results. These include:

* Use a ```-``` to exclude words: if you searched ```mustang -car``` it would
give you results that didn't include car-related information.
* Use the keyword ```search:``` to search within a website: if you searched
```search:stackoverflow.com R cran read csv``` will look for the search words
specifically in stackoverflow.com.
* Use a ```*``` to replace missing words: ```*orientated programming``` would
allow any word to replace the * which helps when you're not sure which word to
use.
* Use double quotation marks ```" "``` to search for words and phrases
specifically.

These are just examples, see more
[here](https://www.lifehack.org/articles/technology/20-tips-use-google-search-efficiently.html).

### Asking in forums for help
Your last resort should be asking online directly (such as posting in
stackoverflow or project github repo) for help. Particularly when starting out,
it is very likely that someone else has encountered the same problem as you in
the past and there is a solution out there.

However, if you really can't find a solution for your error/problem, make sure
your post provides as much information about it as possible:

* version of the program you are installing
* your operating system
* R or Python version
* the exact error message

If you give a developer the exact error message, they know exactly where the
error occurred and much more likely to be able to help you. See this extensive
[guide](https://stackoverflow.com/help/how-to-ask) for further information on
how to ask questions in forums.

# Finding help for a broader problem

## Tutorials
What if you don't have a specific error, maybe it's a data science task that you
don't know how to tackle. The best thing to start with could be a tutorial like
[this](https://swcarpentry.github.io/r-novice-inflammation/11-supp-read-write-csv/)
one. Tutorials are useful because they are usually easy to follow and give
example data. This is an easy way to understand how a function works and how you
would implement it.

## Break down the problem
If you are struggling to find a relative tutorial or you're not sure exactly how
to tackle a problem, the first thing is to break it down as much as possible. It
might seem trivial but when you can visualise the blocks that make up a problem
you can immediately start to see how you would do tackle individual part.

For example: You are given an list and you need to return a list that only
contains the list's even numbers. Theses are some of the questions you should be
asking yourself:

* How to identify an even number? Divide that number by 2 and see if its
remainder is 0.
* What will that list contain? One or more numbers
* What are the data types of the elements in the list? Numbers
* What is the end goal? The goal is to return a list of even numbers. If there
are no even numbers, return an empty list.

By breaking down the problem you will be able to identify individual sections
that you are struggling with and this is essential to find the help you need to
proceed.