---
title: "Intro to R - Chapter Five "
output:
  html_document: 
    theme: united
    highlight: tango
    toc: yes
    toc_float: yes
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("../Images/LATree.PNG"),
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:200px;')
```



<br>

***
# Summary Statistics and Aggregation{-}

***

# Learning Objectives

<br>

**The goal of this chapter is to:**


* Describe numeric and categorical data

* Aggregate and data


***

<br>

# Dataset{-}

We will use the titanic dataset in this chapter with cleaned names via janitor.

```{r message=FALSE, warning=FALSE, include=FALSE}
# Load libraries

library(readr)
library(janitor)
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)


# Read in data

titanic <- readr::read_csv("../Data/titanic.csv", 
                           na=c("", " ", "NULL"))

titanic <- janitor::clean_names(titanic)


```

```{r}
# To display the data

titanic
 
```


***

<br>

# Descriptive Statistics

There are two broad types of data in our `tibble` at the moment that we want to look at - numerical data (i.e. ints and doubles) and text data (i.e. character strings).

In this section, we will explore some basic single variable descriptive statistics.

***

<br>

## Descriptive Statistics for Numerical Data
Let's start with the numerical data, because that's the easiest to work with. The `dplyr::summarise()` function allows us to create one or more scalar variables summarising the variables of an existing `tibble`.  

The useful functions that we can use with `summarise()`:

* `mean()` - The average (arithmetic mean) data value in the given column.
* `median()` - The middle value of an ordered vector of values.
* `sd()` - The standard deviation (spread) of values in the given column.
* `quantile()` - Produces sample quantiles corresponding to the given proportions (between 0 and 1). 
* `min()` - The smallest value in the given column.
* `max()` - The maximum data value recorded.
* `n()` - Counts the number of entries (rows).
* `n_distinct()` - Counts the number of unique entries for a given variable.

<br>

Base R does not provide a function for finding the mode.

We would have to use a package to calculate the mode or write our own function.

The function `mfv()`, for most frequent value, in the `modeest` package can be used to find the statistical mode.

There are plenty more we could use, but this, like the rest of the course, is intended to give you a taste of what's possible. 

Let's look at some examples:
```{r}

# Calculating the mean of the column fare 
# in the titanic dataset

dplyr::summarise(.data = titanic, mean(fare))
```

<br>

>**This highlights an important point, any operation applied to a column containing one or more `NA` values produces `NA`.** 

Most functions in R, for best practice reasons, propagate `NA` values, requiring us to set the common parameter `na.rm` to `TRUE`. 

This will omit any `NA` values in our calculation and produce a more meaningful result, albeit on a subset of observed data. The intention here is that R wants the analyst to know `NA` values are present in the data and deal with them directly. 

```{r}
# Calculating the mean of the column fare 
# in the titanic dataset
# Setting na.rm=TRUE to remove missing values
# from calculation

dplyr::summarise(.data = titanic, mean(fare, na.rm = TRUE))

```

<br>

To assign this a meaningful column name, we'd pass a keyword argument in `summarise()`.

```{r}
# Calculating the mean of the column fare 
# in the titanic dataset
# Setting na.rm=TRUE to remove missing values
# from calculation
# setting the name mean_fare for our result

dplyr::summarise(.data = titanic, mean_fare = mean(fare, na.rm = TRUE))

```

<br>

`summarise()` is effectively computing out statistics and wrapping the result in a `tibble`. 

If we didn't care about the `tibble` output we could simply compute using base R syntax:


```{r}
# Calculating the mean of the column fare 
# in the titanic dataset
# Setting na.rm=TRUE to remove missing values
# from calculation

mean(titanic$fare, na.rm = TRUE)

```

<br>

Like we saw with `mutate()` and other functions, we can compute multiple statistics through `summarise()`.

```{r}
# Calculating muiltiple summary statisitcs within
# the summarise function

dplyr::summarise(.data = titanic, 
                 
                 median_fare = median(fare, na.rm = TRUE),
                 
                 mean_fare = mean(fare, na.rm = TRUE),
                 
                 sd_fare = sd(fare, na.rm = TRUE),
                 
                 median_age = median(age_of_passenger, na.rm = TRUE),
                 
                 mean_age = mean(age_of_passenger, na.rm = TRUE),
                 
                 sd_age = sd(age_of_passenger, na.rm = TRUE))
```


<br>

The structure of our output above is not ideal. 

We could restructure it using the `tidyr` functions: **`pivot_wider()`**, *`separate()`* and **`pivot_longer()`**. This is slightly advanced for what we have covered so far, but examples have been added to the additional materials document for those keen to know.

Functions which return more than one value won't work well with **`summarise()`**. 

Therefore, these functions have to be called on the data directly instead.

```{r}
# Calculating the quartiles from the
# fare column in the titanic dataset

stats::quantile(titanic$fare, na.rm = TRUE)

```

<br>

We can also specify as below,


```{r}
# Calculating the quartiles from the
# fare column in the titanic dataset
# by specifying which points to display

stats::quantile(titanic$fare, 
                probs = c(0.25, 0.5, 0.75, 1),
                na.rm = TRUE)
```

<br>

We can further specify as below,

```{r}
# Calculating the quartiles from the
# fare column in the titanic dataset
# by specifying which points to display

stats::quantile(titanic$fare, 
                probs = c(0, 0.2, 0.4, 0.6, 0.8, 1),
                na.rm = TRUE)

```

***

<br>

## Describing Text (or Categorical) Data 


Simple statistics on qualitative data include:

* `dplyr::distinct()` - Which will list all the unique values in a given column. 

* `dplyr::n_distinct()` - Takes a vector and counts how many unique values there are.

* `dplyr::count()` - Displays the number of occurrences of each unique entry in a given column.


Descriptive statistics below are for qualitative data, they don't work very well if every value in a field is a different piece of text!

```{r}
# Calculating a frequency table for the number
# of passengers and sorting in descending order

dplyr::count(titanic, name_of_passenger, sort = TRUE)

```

<br>

Compared to:
```{r}
# Calculating a frequency table for the number
# of passengers and sorting in descending order

dplyr::count(titanic, embarked, sort = TRUE)

```


<br>

## Exercise {.tabset .tabset-fade}


### **Exercise**{-}

1. How old is the oldest passenger in the dataset?

2. How many men and women are in the dataset?

3. Create a new column called 'std_fare' which is the 'fare' minus the mean fare, divided by the standard deviation.

4. Calculate the number of children in second class.

***

<br>

### **Hint**{-}

1. How old is the oldest passenger in the dataset?

```{r, eval=FALSE}

# max gives you the highest value

oldest_passenger <- max(data)

```

<br>

2. How many men and women are in the dataset?

```{r, eval=FALSE}

# Count gives a frequency

number_of_men_and_women <- dplyr::count(data, columns)

```

3. Create a new column called `'std_fare'` which is the `'fare'` minus the mean fare, divided by the standard deviation.

You will need to use you data manipulation skills from previous chapter.

```{r, eval=FALSE}
# Mutate lets you add new variables to the dataset

standard_fare <- dplyr::mutate(data, 
                               name of new column = contents of new column)

```

***

<br>

### **Show Answer**{-}

1. How old is the oldest passenger in the dataset?

```{r}
# Calculating the highest values in the 
# age_of_passenger column

oldest_passenger <- max(titanic$age_of_passenger, 
                              na.rm = TRUE)

```

2. How many men and women are in the dataset?

```{r}
# Calculating a frequency count of the
# sex_of_passenger column

number_of_men_and_women <- dplyr::count(titanic, 
                                        sex_of_passenger)

```

3. Create a new column called 'std_fare' which is the 'fare' minus the mean fare, divided by the standard deviation.

```{r}

# Using mutate to add a new column on the titanic dataset

standard_fare <- dplyr::mutate(.data=titanic, 
                               std_fare = 
                                 (fare - mean(fare, na.rm = TRUE)) / sd(fare, na.rm = TRUE))

```

***

<br>

# Aggregation


Aggregation means grouping data together by a particular grouping variable and producing a summary of one or more columns for that grouping variable.

We'll use the **`dplyr::group_by()`** function.

This function can be really useful, especially when your data are disaggregate - e.g. data about individual units of people or things. 

**`group_by()`** allows you to aggregate by a categorical variable and then summarise numerical data into a new `tibble`. It works on a principle known as 'split-summarise-combine':

* Split - a `tibble` is divided into a set of smaller `tibbles` based on the grouping variable.

* Summarise - an aggregation is applied to each of the groups to create a single row for each group in the original `tibble`.

* Combine - bring together the aggregated `tibble` rows into a final new `tibble`.

Let's walk through what that might look like for the `titanic` `tibble`:

* Firstly, we decide to **split** the data by the `'pclass'`. This divides the `titanic` `tibble` into effectively three separate `tibbles`, one for first, one for second and one for third class.

* Secondly, we **summarise** the `tibble` using a chosen aggregation function. You can either produce an aggregate statistic for all columns, or you can select specific columns on which to do the aggregation. 

* Finally, we get a **combined** `tibble` that contains the new aggregate statistics.

Let's look at that in code:

```{r}

# Grouped tibble by pclass

by_class <- dplyr::group_by(titanic, pclass)

```

<br>

```{r echo=FALSE}

by_class %>% 
  kable() %>% 
   kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100", height = "500px")
 
```

<br>

Now let's find out the mean fare paid by each class

```{r}
# calculating the mean from the fare column in the grouped 
# dataset

dplyr::summarise(by_class, 
                 mean_fare = mean(fare, na.rm = TRUE))
```

<br>

**`group_by()`** is a powerful tool, particularly when you are working with any kind of hierarchical data where you might want to know something aggregate about the groups within the data, for instance:

* individuals nested in households.

* employees nested in firms.

* small area geographies (e.g. wards, output areas, postcodes etc.) nested in larger geographies (e.g. districts, counties etc.)

* countries nested in supra-national entities or, demographic, cultural and socio-economic classes.

* individuals by age, sex, ethnicity, religion etc.

* employees by grade or occupational social class.




>**We can also aggregate according to more complicated groupings**

Here we are grouping by 2 columns `"pclass"` and `"embarked"`.

```{r}
# Group by pclass and embarked. Then find the average fare 
# paid by each group.

by_class_embarked <- dplyr::group_by(titanic,
                                     pclass, embarked) %>%
                      dplyr::summarise(mean_fare = mean(fare, na.rm = TRUE))


# To display the data

by_class_embarked

```

```{r echo=FALSE, eval=FALSE}

by_class_embarked %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100", height = "500px")
 
 
```

<br>

Suppose with the above, the `NA` values bother us. To remove them we can add some code to our pipeline to handle this. We can use the **`drop_na()`** function from the tidyr package.

Here I have added some extra summary statistics as well, i.e standard deviation and a count.

```{r}
# Calculating the median fare
# dropping missing values using drop_na from tidyr package
# grouping by pclass and emabrked
# summarising by shwoing the median fare and standard deviation
# and a frequency count

median_fare <- titanic %>%
  
  drop_na(embarked) %>%
  
  group_by(pclass, embarked) %>%
  
  summarise(median_fare= median(fare, na.rm = TRUE),
            
            sd_fare = sd(fare, na.rm = TRUE),
            
            count = n())

#To display the data

median_fare


```

<br>


I plan on presenting this data, so I want something more meaningful in place than "C", "Q" and "S".

```{r}

# Calculating the median fare
# dropping missing values using drop_na from tidyr package
# using mutate to recode the variables "C","Q" and "S"
# grouping by pclass and emabrked
# summarising by shwoing the median fare and 
# standard deviation and a frequency count

median_fare <- titanic %>%
  
  drop_na(embarked) %>% 
  
  mutate(embarked = recode(embarked, 
                           "C" = "Cherbourg", 
                           "Q" = "Queenstown", 
                           "S" = "Southampton")) %>%
  
  group_by(pclass, embarked) %>%
  
  summarise(
    
    median_fare= median(fare, na.rm = TRUE),
            
    sd_fare = sd(fare, na.rm = TRUE),
            
    count = n()
    
  )

#To display the data

median_fare

```

<br>

We can simply visualise our data in a chart. (We are not covering how to plot here.)

<br>

```{r, echo=FALSE}

ggplot2::ggplot(median_fare,
       aes(x = pclass,
           y = count))+
  geom_col(position="dodge",aes(fill=embarked))+theme_grey()
```


<br>

## Exercise {.tabset .tabset-fade}


### **Exercise**{-}

1. What is the average fare paid by men and women?


2. What is the median fare paid by men and women in each different class?


***

<br>

### **Hint**{-}

1. What is the average fare paid by men and women?

```{r, eval=FALSE}
# Calculating the average fare
# Grouping by sex_of_passenger
# summary showing the mean fare

average_fare <- titanic %>% 
  
  group_by() %>% 
  
  summarise()

```

<br>

2. What is the median fare paid by men and women in each different class?

```{r}
# Calculating the median fare
# Grouping by sex_of_passenger and pclass
# summary showing the mean fare

median_fare <- titanic %>%
  
  group_by() %>%
  
  summarise()

```


***

<br>

### **Show Answer**{-}

1. What is the average fare paid by men and women?

```{r}
# Calculating the average fare
# Grouping by sex_of_passenger
# summary showing the mean fare

average_fare <- titanic %>% 
  
  group_by(sex_of_passenger) %>% 
  
  summarise(mean_fare = mean(fare, na.rm = TRUE))

#To display the data

average_fare
```

<br>

2. What is the median fare paid by men and women in each different class?

```{r}
# Calculating the median fare
# Grouping by sex_of_passenger and pclass
# summary showing the mean fare
median_fare <- titanic %>%
  
  group_by(sex_of_passenger, pclass) %>%
  
  summarise(median_fare = median(fare, na.rm = TRUE))

#To display the data
median_fare
```


***

<br>


# Summary - Summary Statistics and Aggregation{-}

<br>

## Reminder of Learning Objectives{-}

**You should now be able to**


* Understand how to obtain summary statistics.

* Be able to use R functions to obtain descriptive statistics.

* Be able to aggregate data in a data frame.

***

<br>

# Next Chapter

***

>**In the next chapter we will look at Control Flow and Functions.**

***


